Reproducing ``Fluctuation Domains''

### 2020-02-18 rescience/manuscript.tex
Fix some typos

In this analyis I seek to reproduce the numerical results of (Boettiger,
Dushoff, and Weitz 2010), which focuses on the evolutionary trajectory
of species trait under selection, as predicted by the so-called
``Canonical Equation'' of adaptive dynamics (Dieckmann and Law 1996).
Theoretical work I present in the paper predicts the emergence of a
`flutation enhancement' regime far from an evolutionary stable point,
and suggests that the evolution of traits which start in this distant
region would deviate substantially from the expected evolutionary path
predicted by the canonical equation. In particular, the distribution of
such trajectories could be bimodal, making the ``canonical'' path far
from the most likely path. The details of this distribution cannot be
solved for analytically however, so in this paper I resorted to an
replicate exact stochastic simulations of the evolutionary model using a
Gillespie algorithm (Gillespie 1977).

@@

The published paper itself states that the code is available at the
personal website of one of my co-authors, with URL
\url{http://ecotheory.biology.gatech.edu/downloads/}. Fortunately, this
URL still resolves (my colleague is still at the same institution), and
does indeed contain a downloadable zip file with a copy of the code. In
2012, I had uploaded a copy of this code as a supplement to Dryad, but
both on my CV a the time and on my colleague's website we had recorded
only the Dryad handle URL,
\url{http://datadryad.org/handle/10255/dryad.37625}, which no longer
resolves. That handle can be resolved by the WayBackMachine, which
doesn't contain the code archive but does reveal the DOI,
\url{http://dx.doi.org/10.5061/dryad.j8n0p7vc}, which successfully
resolves to the Dryad copy. The code can also be found by searching the
Dryad web interface for my surname (though it is listed as ``data'' and
not ``software.'' In 2017, I had also imported the GitHub copy into the
Zenodo data repository where it was again assigned a DOI and is
identified as software in the metadata. I note that the package was
never submitted to the Central R Archive Network, CRAN.

@@

\begin{quote}
In the spirit of scientifically reproducible research(Gentleman and
Lang, 2007; Schwabet al., 2000; Stodden, 2009), we freely provide all
the source code required to replicate the simulations and figuress hown
in the text. Though the numerical simulations are written in C for
computational efficiency, we provide a user interface and documentation
by releasing all the code, figures, text, and examples as a software
package for the widely used and freely available R statistical computing
language. This package can be downloaded from
\url{http://ecotheory.biology.gatech.edu/downloads/}.
\end{quote}

@@

The same stategy worked on MacOSX (Catalina) with \texttt{gsl} installed
via Homebrew. I was able to update the original autoconf script by
recalling that it had been adapted from another package, the
\texttt{gsl} R package
\url{https://cran.r-project.org/web/packages/gsl/}, which also binds
against the GSL library. Working from the autoconf files provided there,
I could successfully get the package to configure and compile. This
highlights a potential risk in changes that are intended to make the
package more portable but also add complexity. More generally, while the
package has no dependency on external R packages (beyond those default
packages provided as part of base R: \texttt{stats}, \texttt{utils},
\texttt{graphics}, and \texttt{grDevices}, this issue illustrates the
familiar reproducibility challenge that is created by changing versions
of dependencies and system libraries (GSL in this case). System
libraries against which a package is compiled are particularly
pernicious, because versioning of these components is both platform
dependent and separate from the R package manager process, and thus
cannot be addressed merely with package version tracking in tools such
as \texttt{packrat} (Ushey et al. 2018) or \texttt{renv} (Ushey 2019).
Today, containerized approaches such a Docker (see Boettiger 2015;
Boettiger and Eddelbuettel 2017) provide a solution to this
system-dependency issue that was not available a decade ago.

The second challenge was in figuring out just what commands had to be
run to duplicate the figures. As mentioned above, the GitHub version of
the source-code did not have the manuscript source, and even the version
that did have the mansucript source did not write the manuscript itself
as a dynamic document (e.g.~in Sweave (Xie 2015)). In that version,
files \texttt{fig1.R} and \texttt{fig2.R} are were included in the same
subdirectory as the \texttt{pdf} figures referenced by the \texttt{tex}
version of the manuscript, but \texttt{fig2.R}, but these files
contained only the code required to actually generate the figures
themselves once the result data had been pre-computed and saved as
\texttt{.txt} files. That result data files had also not been archived
seperately. Fortunately, re-reading my own documentation, I was able to
deduce that these text files were generated by running the function
\texttt{logistic()} with the parameter settings given in the caption of
Fig 2 in the manuscript.

@@

Running the \texttt{logistic()} function provided in the code, I was
able to replicate the key results of the paper. A few qualifications
here are necessary. The paper consists of two core results -- a
theoretical (i.e.~derived analytically without computational component)
which predicts the existance of `fluctuation enhancement regimes' which
can create substantial deviations from the dynamics predicted by the
accepted theory at the time (that is, from the so-called canonical
equation of adaptive dynamics), and a second result which relies on a
computationally intensive stochastic ensemble simulation to illustrate
these deviations arising for dynamics that begin in the ``fluctation
enhancement'' regime. These results are summarized by figures 1 and 2 of
the original paper, which I reproduce here.

@@

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/fig1-1.pdf}
\caption{The three panels of figure 1, showing examples of evolutionary
landscapes and the corresponding fluctation domains for three classic
models (logistic, chemostat, branching model). This figure is largely
theoretical, and so easily replicates Figure 1 of the original paper.
The first panel showing the logistic model depicts the fluctation
regimes for the model parameters used in the numerical analysis.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/fig1-2.pdf}
\caption{The three panels of figure 1, showing examples of evolutionary
landscapes and the corresponding fluctation domains for three classic
models (logistic, chemostat, branching model). This figure is largely
theoretical, and so easily replicates Figure 1 of the original paper.
The first panel showing the logistic model depicts the fluctation
regimes for the model parameters used in the numerical analysis.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/fig1-3.pdf}
\caption{The three panels of figure 1, showing examples of evolutionary
landscapes and the corresponding fluctation domains for three classic
models (logistic, chemostat, branching model). This figure is largely
theoretical, and so easily replicates Figure 1 of the original paper.
The first panel showing the logistic model depicts the fluctation
regimes for the model parameters used in the numerical analysis.}
\end{figure}

@@

A dynamic document using RMarkdown (Xie, Allaire, and Grolemund 2018)
for reproducing the present mansucript can be found in the
\texttt{rescience} sub-directory which I have now added to the GitHub
repository housing the original R package,
\url{https://github.com/cboettig/fluctuationDomains}, git tag
\texttt{rescience-submission}, now archived with DOI
\url{https://doi.org/xxxx}.

>>>>

In this analysis I seek to reproduce the numerical results of (Boettiger,
Dushoff, and Weitz 2010), which focuses on the evolutionary trajectory
of species trait under selection, as predicted by the so-called
``Canonical Equation'' of adaptive dynamics (Dieckmann and Law 1996).
Theoretical work I present in the paper predicts the emergence of a
`flutation enhancement' regime far from an evolutionary stable point,
and suggests that the evolution of traits which start in this distant
region would deviate substantially from the expected evolutionary path
predicted by the canonical equation. In particular, the distribution of
such trajectories could be bimodal, making the ``canonical'' path far
from the most likely path. The details of this distribution cannot be
solved for analytically however, so in this paper I resorted to an
replicate exact stochastic simulations of the evolutionary model using a
Gillespie algorithm (Gillespie 1977).

@@

The published paper itself states that the code is available at the
personal website of one of my co-authors, with URL
\url{http://ecotheory.biology.gatech.edu/downloads/}. Fortunately, this
URL still resolves (my colleague is still at the same institution), and
does indeed contain a downloadable zip file with a copy of the code. In
2012, I had uploaded a copy of this code as a supplement to Dryad, but
both on my CV at the time and on my colleague's website we had recorded
only the Dryad handle URL,
\url{http://datadryad.org/handle/10255/dryad.37625}, which no longer
resolves. That handle can be resolved by the WayBackMachine, which
doesn't contain the code archive but does reveal the DOI,
\url{http://dx.doi.org/10.5061/dryad.j8n0p7vc}, which successfully
resolves to the Dryad copy. The code can also be found by searching the
Dryad web interface for my surname (though it is listed as ``data'' and
not ``software.'' In 2017, I had also imported the GitHub copy into the
Zenodo data repository where it was again assigned a DOI and is
identified as software in the metadata. I note that the package was
never submitted to the Central R Archive Network, CRAN.

@@

\begin{quote}
In the spirit of scientifically reproducible research(Gentleman and
Lang, 2007; Schwabet al., 2000; Stodden, 2009), we freely provide all
the source code required to replicate the simulations and figures shown
in the text. Though the numerical simulations are written in C for
computational efficiency, we provide a user interface and documentation
by releasing all the code, figures, text, and examples as a software
package for the widely used and freely available R statistical computing
language. This package can be downloaded from
\url{http://ecotheory.biology.gatech.edu/downloads/}.
\end{quote}

@@

The same strategy worked on MacOSX (Catalina) with \texttt{gsl} installed
via Homebrew. I was able to update the original autoconf script by
recalling that it had been adapted from another package, the
\texttt{gsl} R package
\url{https://cran.r-project.org/web/packages/gsl/}, which also binds
against the GSL library. Working from the autoconf files provided there,
I could successfully get the package to configure and compile. This
highlights a potential risk in changes that are intended to make the
package more portable but also add complexity. More generally, while the
package has no dependency on external R packages (beyond those default
packages provided as part of base R: \texttt{stats}, \texttt{utils},
\texttt{graphics}, and \texttt{grDevices}, this issue illustrates the
familiar reproducibility challenge that is created by changing versions
of dependencies and system libraries (GSL in this case). System
libraries against which a package is compiled are particularly
pernicious, because versioning of these components is both platform
dependent and separate from the R package manager process, and thus
cannot be addressed merely with package version tracking in tools such
as \texttt{packrat} (Ushey et al. 2018) or \texttt{renv} (Ushey 2019).
Today, containerized approaches such as Docker (see Boettiger 2015;
Boettiger and Eddelbuettel 2017) provide a solution to this
system-dependency issue that was not available a decade ago.

The second challenge was in figuring out just what commands had to be
run to duplicate the figures. As mentioned above, the GitHub version of
the source-code did not have the manuscript source, and even the version
that did have the manuscript source did not write the manuscript itself
as a dynamic document (e.g.~in Sweave (Xie 2015)). In that version,
files \texttt{fig1.R} and \texttt{fig2.R} are were included in the same
subdirectory as the \texttt{pdf} figures referenced by the \texttt{tex}
version of the manuscript, but these files
contained only the code required to actually generate the figures
themselves once the result data had been pre-computed and saved as
\texttt{.txt} files. The result data files had also not been archived
separately. Fortunately, re-reading my own documentation, I was able to
deduce that these text files were generated by running the function
\texttt{logistic()} with the parameter settings given in the caption of
Fig 2 in the manuscript.

@@

Running the \texttt{logistic()} function provided in the code, I was
able to replicate the key results of the paper. A few qualifications
here are necessary. The paper consists of two core results -- a
theoretical (i.e.~derived analytically without computational component)
which predicts the existence of `fluctuation enhancement regimes' which
can create substantial deviations from the dynamics predicted by the
accepted theory at the time (that is, from the so-called canonical
equation of adaptive dynamics), and a second result which relies on a
computationally intensive stochastic ensemble simulation to illustrate
these deviations arising for dynamics that begin in the ``fluctuation
enhancement'' regime. These results are summarized by figures 1 and 2 of
the original paper, which I reproduce here.

@@

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/fig1-1.pdf}
\caption{The three panels of figure 1, showing examples of evolutionary
landscapes and the corresponding fluctuation domains for three classic
models (logistic, chemostat, branching model). This figure is largely
theoretical, and so easily replicates Figure 1 of the original paper.
The first panel showing the logistic model depicts the fluctuation
regimes for the model parameters used in the numerical analysis.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/fig1-2.pdf}
\caption{The three panels of figure 1, showing examples of evolutionary
landscapes and the corresponding fluctuation domains for three classic
models (logistic, chemostat, branching model). This figure is largely
theoretical, and so easily replicates Figure 1 of the original paper.
The first panel showing the logistic model depicts the fluctuation
regimes for the model parameters used in the numerical analysis.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/fig1-3.pdf}
\caption{The three panels of figure 1, showing examples of evolutionary
landscapes and the corresponding fluctuation domains for three classic
models (logistic, chemostat, branching model). This figure is largely
theoretical, and so easily replicates Figure 1 of the original paper.
The first panel showing the logistic model depicts the fluctuation
regimes for the model parameters used in the numerical analysis.}
\end{figure}

@@

A dynamic document using RMarkdown (Xie, Allaire, and Grolemund 2018)
for reproducing the present manuscript can be found in the
\texttt{rescience} sub-directory which I have now added to the GitHub
repository housing the original R package,
\url{https://github.com/cboettig/fluctuationDomains}, git tag
\texttt{rescience-submission}, now archived with DOI
\url{https://doi.org/xxxx}.
